Functional trait data tell new story about biodiversity change

### 2021-02-24 paper/manuscript.Rmd
Revamp methods, refine intro.

journal: "Ecography"
date: "`r Sys.Date()`"
bibliography: refs.bib
layout: 3p
header-includes:
   - \usepackage{lineno}
   - \linenumbers
output: 
  rticles::elsevier_article:
    includes:
      in_header: preamble.tex
editor_options: 
  markdown: 
    wrap: 72
---

@@

Ecological communities are experiencing unprecedented change as a result
of anthropogenic pressures such as climate change, land use change, and
invasive species.

#### Paragraph 1 - communities are changing, we need to know how and what that means for ecosystem properties

-   communities are changing as a result of a whole bunch of different
    pressures

-   a few well documented phenomena: accelerating extinction rates, what
    else?

-   it's essential for conservation and ecology to understand what these
    changes mean for ecosystem properties like stability, function, etc

#### Paragraph 2 - species diversity is the go to - it's not working

-   traditionally the most common method of capturing change in
    communities is biodiversity

-   despite other well documented changes, species diversity doesn't
    seem to capture local community-level changes
    ([@dornelasAssemblageTimeSeries2014],
    [@suppSpecieslevelCommunitylevelResponses2014],
    [@vellendGlobalMetaanalysisReveals2013]), though there is evidence
    of a significant increase in turnover

-   We need to look beyond species and their abundances to understand
    the kinds of changes happening in communities

#### Paragraph 3 - functional diversity is the answer!

-   Functional diversity provides a potentially powerful tool for
    examining biodiversity changes because it captures characteristics
    beyond species identity.

-   Powerful functional lens ([@mcgillRebuildingCommunityEcology2006])

-   Functional diversity can tells us a story independent of species
    richness, as huge functional changes can occur while species
    richness (or even diversity!) stays the same. This especially gives
    insight into the turnover we know is occurring

-   We examine 100's of mammal, bird, and amphibian timeseries from the
    Biotime database paired with functional trait data to assess overall
    trends in functional change and identify where functional diversity
    tells us a story beyond species richness

@@

Biodiversity timeseries were obtained from the Biotime database, a
global repository of assemblage timeseries with 386 individual studies
from multiple taxa. Following best practices for the dataset
[@blowes2019] to account for differences in scale between studies, we
aggregated based on a grid system:

*annotation of* `01_biotime_grid.R`

-   label single location (SL) or multiple location (ML), ML with an
    extent less than ((SL extent mean + SL extent SD) are assigned to be
    SL

    -   SL coords are the central lat and long, ML coords are the center
        of convex hull of locations within a rarefyID (grid cell)

-   exclude studies with one year of data

-   create global grid with resolution based on the cutoff study area
    size (SL extent mean + SL extent SD)

-   collate observations for the same study within cells/years

-   remove years (for count) or rarefyID's (for abundance and biomass)
    that do not meet coverage requirement of

Timeseries were paired with their functional trait data based on the
predominant taxa in the timeseries. Due to limitations in trait data and
processing feasability, we included only mammal, bird, and amphibian
timeseries. Trait data were obtained from Elton Trait Database (mammals
and birds; [@wilman2014]) and Amphibio (amphibians, [@oliveira2017]).

-   something here describing where the data in those databases comes
    from, that they're species-level means, etc**;**

-   also more description of how biotime is structured (studies, some
    with plots, etc)

In order to ensure taxonomic consistency across datasets, biotime
species were paired with trait data based on their species identifier
from the Integrated Taxonomic Information System database (citation),
obtained through the taxadb R package
([@normanTaxadbHighperformanceLocal2020]). If more than one species in
the assemblage data resolved to the same identifier, observations were
considered the same species. For trait data, traits for all species of
the same identifier were averaged. Only studies with at least 85% trait
coverage were included. In order to have a sufficient number of
observation years to observe a trend and a sufficient number of species
to calculate functional diversity metrics, studies with fewer than 4
years and years with fewer than 4 species observed were also excluded.

Functional diversity metrics were calculated for total yearly abundance.
For many studies there were a different number of samples within each
year. To adjust for this inconsistency in sampling effort we used
sample-based rarefaction by bootstrap resampling within year for each
timeseries based on the smallest number of samples in a year for that
timeseries.

-   raw abundance counts within a year for each plot

@@

We calculated yearly metrics of functional diversity and turnover for
each plot within each study. All available trait data for each study was
included in the trait matrix with the exception of traits that were the
same value for all observed species in the study. All continuous traits
were z-score scaled to give each trait equal weight in the trait space
([@lepsQuantifyingInterpretingFunctional2006],
[@schleuterUserGuideFunctional2010]).

We used the function dbFD from the FD package (citation) to calculated
PCA-scaled trait matrices for each species in a timeseries and
functional diversity (FD), functional evenness (FEve), functional
diveregence (FDiv), and functional dispersion (FDis) based on those
trait matrices. The number of trait axes was limited to the maximum
number of traits that fulfills the criteria $s >= s^t$, where $s$ is the
number of species and \$t\$ is the number of traits. This limitation
allows for a sufficient number of axes to capture the trait space while
maintaining computational feasibility for metric calculations. Metrics
incorporated weighting based on either species abundance or biomass
based on the methods for each study. The PCA-scaled trait matrices were
then used as input to the BAT package....

-   needed euclidian distance matrix - used calliez correction because
    it was a sufficient correction for all plots even though it led to a
    marginal decrease in "quality" for the corrected hull

-   null model stuff via [@swensonBiogeographyFilteringWoody2012] - for
    years where all species in the species pool were observed the null
    model sd is zero for functional richness, since FRic does not
    incorporate abundance changes. - drop species with trait match

@@

>>>>

journal: "Ecography"
date: "`r Sys.Date()`"
bibliography: refs.bib
layout: 3p
header-includes:
   - \usepackage{lineno}
   - \linenumbers
output: word_document
  #rticles::elsevier_article:
   # includes:
     # in_header: preamble.tex
editor_options: 
  markdown: 
    wrap: 72
---

@@

Ecological communities are experiencing unprecedented change as a result
of anthropogenic pressures such as climate change, land use change, and
invasive species. Impacts of these pressures are well documented on a
global scale with an accelerating global extinction rate (citation) and
fundamental changes in some of the most well-studied systems (e.g. coral
bleaching). At the local scale however, species diversity tells a
different story. Recent synthesis of local trends in biodiversity over
time have shown a decoupling of community composition from richness
[@brose2016], evidence of community-level regulation of richness
[@gotelli2017], and no net change in local biodiversity despite
widespread environmental change and ongoing species turnover
[@brown2001; @dornelas2014; @vellend2013; , @vellend2017]. While
communities are clearly changing, our most common species-based
approaches do not fully capture the nature of that change.

Functional diversity offers a potentially powerful alternative for
detecting and describing community change by providing a mechanistic
link between the organisms in a system and the processes they perform
[@mcgill2006]. By summarizing functional traits rather than species,
functional diversity metrics capture the disproportionate impact of
losses or gains of functionally unique species. Functional diversity
metrics are particularly suited for understanding local-scale
biodiversity patterns as they can detect shifts in community structure
independent of species change.

<!-- Functional diversity also sheds light on critical ecological processes. -->

Functional diversity also sheds further light on critical ecosystem and
community processes. For example, there is increasing evidence that
functional diversity is a better predictor of ecosystem function than
species-based metrics [@cadotte2011; @gagic2015], and that different
facets of functional diversity also play a role in maintaining ecosystem
stability [\@craven2018; @morin2014]. Further, biogeographic assessments
of functional diversity give foundational insight into how community
assembly plays out across systems at a standardized scale [@violle2014;
@weiher2011].

<!-- What can we say about how global change is impacting:stability, resilience, function, assembly, ecosystem processes (and ecosystem services) - need a functional grounding for all those concepts. Link specific patterns in FD metrics to impacts on those concepts. -->

Assessments of broad-scale temporal change in functional diversity have
previously been limited by a lack of functional trait data. The majority
of work has therefore focused largely on system-specific studies with
traits collected *in situ*. Ongoing efforts to assemble functional
traits for a variety of taxa have made synthesis of existing community
assemblage data and functional traits possible for the first time,
providing initial looks at how functional diversity changes on a
broad-scale [@trindade-santos2020; @jarzyna2016; @barnagaud2017].
<!-- so what do these studies say? does that go here? --> However to
date there has been no comprehensive assessment of functional change
through time on a broad geographic and taxonomic extent.\

Here we perform the first cross-taxa and cross-system assessment of
functional diversity change through time. We assess 100's of mammal,
bird, and amphibian functional diversity timeseries to determine how
functional diversity change compares to species change. We also identify
3 distinct types of functional change common across communities and
their implications for ecological processes.

@@

We obtained mammal, bird, and amphibian timeseries from the Biotime
database, a global repository of high quality assemblage timeseries. All
studies included in the database follow consistent sampling protocols
and represent full assemblages rather than population-level data
[@dornelas2018]. Following best practices for the database [@blowes2019]
and to maintain a consistently defined local scale, studies with
multiple sample locations were split into separate time series based a
global grid wherein all samples within a single grid cell were
considered a single time series. The size of the cell was determined
based on the sample extent of the single location studies (see
[@dornelas2018] for details on how sample extents were defined), with
the area of each cell set to one standard deviation away from the mean
of the single extent locations.

We gathered trait data from the Elton Trait Database (mammals and birds;
[@wilman2014]) and Amphibio (amphibians, [@oliveira2017]). These
databases include species-level means for traits that best define
species' function in the community including body size, diet, and
behavioral characteristics <!-- what's in the amphibian database? -->.
For the full list of traits included in the analysis see the
supplemental materials.

In order to ensure taxonomic consistency across datasets, biotime
species were paired with trait data based on their species identifier
from the Integrated Taxonomic Information System database (citation),
obtained through the taxadb R package ([@norman2020]). If more than one
species in the assemblage data resolved to the same identifier,
observations were considered the same species. For trait data, traits
for all species of the same identifier were averaged. Only studies with
at least 85% trait coverage were included. In order to have a sufficient
number of species to calculate functional diversity metrics years with
fewer than 4 species observed were excluded.

Many studies had a variable number of samples within years. To adjust
for this sampling effort inconsistency we used sample-based rarefaction
by bootstrap resampling within years for each timeseries based on the
smallest number of samples in a year for that timeseries.

@@

We calculated yearly metrics of functional diversity and turnover for
each timeseries within each study. All available trait data for each
study was included in the trait matrix with the exception of traits that
were the same value for all observed species in the study. All
continuous traits were z-score scaled to give each trait equal weight in
the trait space ([@leps2006; @schleuter2010]). We also calculated basic
species richness, *S.*

We used the function dbFD from the FD package (citation) to calculated
PCA-scaled trait matrices for each species in a timeseries and
functional diversity (FD), functional evenness (FEve), functional
diveregence (FDiv), and functional dispersion (FDis) based on those
trait matrices. The number of trait axes was limited to the maximum
number of traits that fulfills the criteria $s >= s^t$, where $s$ is the
number of species and \$t\$ is the number of traits. This limitation
allows for a sufficient number of axes to capture the trait space while
maintaining computational feasibility for metric calculations. Metrics
incorporated weighting based on either species abundance or biomass
based on the methods for each study.

@@

We modeled each diversity metric using a Generalized Linear Mixed
Effects model with year as the predictor and mixed effects of timeseries
nested within study. <!-- more to come! -->
